---
title: "rent_growth_v_pre_pandemic_trend"
output: html_document
date: "2026-01-22"
---

```{r}
library(bigrquery)
library(tidyverse)
library(DBI)
```

```{r}
con <- dbConnect(
  bigrquery::bigquery(),
  project = "alist-production-scratch",
  dataset = "salviati",
  billing = "alist-production-scratch"
)
```

```{r}
index_data <- dbGetQuery(con,
  "
  select 
    i.location_id, 
    l.display_name,
    case when i.location_id = 0 then 'National'
         when i.location_id is not null and l.display_name is null then 'Metro'
         else l.type end as location_type,
    case when i.location_id = 0 then 'United States'
         when i.location_id is not null and l.display_name is null then xwalk2.cbsa_name
         else l.display_name end as location_name,
    i.year, 
    i.month, 
    i.population,
    i.moving_avg_price_index,
    i.price_all,
    i.price_1br,
    i.price_2br,
    i.transaction_count,
    xwalk.state_fips_code,
    xwalk.state_name as state,
    xwalk.county_fips_code,
    xwalk.county_name as county,
    xwalk.place_fips_code,
    xwalk.place_name as place,
    xwalk.cbsa_fips_code,
    xwalk.cbsa_name as metro
  from `alist-production.analytics.rent_report_price_indexes` i   
  left join `alist-production.api.locations` l
    on i.location_id = l.id
  left join `alist-production-scratch.rwarnock.al_to_fips_mapping` xwalk
    on i.location_id = xwalk.id
  left join (select distinct cbsa_fips_code, cbsa_name from `alist-production-scratch.rwarnock.al_to_fips_mapping`) xwalk2
    on cast(i.location_id as string) = xwalk2.cbsa_fips_code
  where l.type is null or l.type in ('State', 'County', 'City');
  ") 
```

```{r}
start_date <- ymd("2020-03-01")
base_date  <- start_date %m-% months(1)  # Typically 2020-02-01

prepan_year_min <- 2017L
prepan_year_max <- 2019L
```

Build a working frame with a proper date column.

```{r}
df_locs <- index_data %>%
  mutate(date = ymd(sprintf("%04d-%02d-01", year, month)))
```

Identify the locations to keep: National, the 50 largest metros, all states, and the
100 largest cities. "Largest" is by max observed population, and we require full
historical coverage (same month count as the national series) so partial-history
locations don't skew the 2017-2019 seasonal growth averages.

```{r}
month_ct_natl <- df_locs %>%
  filter(location_id == 0) %>%
  count() %>%
  pull()

select_top_n_locations <- function(df, type, n = Inf) {
  df %>%
    filter(location_type == type, !is.na(moving_avg_price_index)) %>%
    group_by(location_id, population) %>%
    summarise(month_ct = n(), .groups = "drop") %>%
    filter(month_ct == month_ct_natl) %>%
    group_by(location_id) %>%
    summarise(population = max(population, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(population)) %>%
    { if (is.finite(n)) slice_head(., n = n) else . } %>%
    pull(location_id)
}

top_100_metros <- select_top_n_locations(df_locs, "Metro", 100)
top_100_cities <- select_top_n_locations(df_locs, "City", 100)
all_states     <- select_top_n_locations(df_locs, "State")   # no cap: keep every state

locations_keep <- c(0, top_100_metros, all_states, top_100_cities)

df_keep <- df_locs %>%
  filter(location_id %in% locations_keep) %>%
  select(
    location_id, location_type, location_name, year, month, date,
    population, price_all, moving_avg_price_index
  )

# Carry a single population figure per location forward (most recent non-missing
# observation), for use in downstream visualizations (e.g. bubble-map sizing).
pop_lookup <- df_keep %>%
  filter(!is.na(population)) %>%
  group_by(location_id) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  select(location_id, population_latest = population)

# Quick check: should be 1 (national) + up to 50 metros + all states + up to 100 cities
temp <- df_keep %>%
  distinct(location_id) %>%
  summarise(n_locations = n())
```

Compute actual MoM growth rates

MoM growth is calculated using moving_avg_price_index.
```{r}
df_mom <- df_keep %>%
  arrange(location_id, date) %>%
  group_by(location_id) %>%
  mutate(
    price_all_actual = price_all,
    mom_chg_actual = (moving_avg_price_index / lag(moving_avg_price_index)) - 1
  ) %>%
  ungroup()

```

Compute average MoM growth during 2017-2019, both for each location individually and
for the national series (the latter gets reused for every location's "national
average" extrapolation line below).
```{r}
avg_prepan_growth_monthly <- df_mom %>%
  filter(year >= prepan_year_min, year <= prepan_year_max) %>%
  group_by(location_id, month) %>%
  summarise(
    avg_mom_growth_2017_2019_m = mean(mom_chg_actual, na.rm = TRUE),
    .groups = "drop"
  )

national_avg_growth_monthly <- avg_prepan_growth_monthly %>%
  filter(location_id == 0) %>%
  select(month, avg_mom_growth_natl_m = avg_mom_growth_2017_2019_m)

# Quick peek: national seasonal pattern
avg_prepan_growth_monthly %>%
  filter(location_id == 0) %>%
  arrange(month)
```

Construct the counterfactual pre-pandemic trend(s)

Method:

Use actual prices through February 2020 (or last available pre-March observation).

From March 2020 onward, apply a constant MoM growth rate. Two variants are built for
every non-national location:

  - `price_all_prepan_trnd` / `mom_chg_prepan_trnd`: uses that location's OWN
    2017-2019 average seasonal growth rate (location-specific trend).
  - `price_all_prepan_trnd_natl` / `mom_chg_prepan_trnd_natl`: uses the NATIONAL
    2017-2019 average seasonal growth rate applied to that location's own anchor
    price (national-average trend). For the national series itself these two are
    identical by construction.

Pre-period values match actuals exactly in both variants.

```{r}
final_df <- df_mom %>%
  left_join(avg_prepan_growth_monthly, by = c("location_id", "month")) %>%
  left_join(national_avg_growth_monthly, by = "month") %>%
  group_by(location_id) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    is_post = date >= start_date,

    # Anchor price: prefer Feb 2020, else last observed pre-March 2020
    base_price = {
      p_feb <- price_all_actual[date == base_date]
      if (length(p_feb) == 1 && !is.na(p_feb)) {
        p_feb
      } else {
        p_last <- price_all_actual[date < start_date]
        if (length(p_last) == 0) NA_real_ else tail(p_last, 1)
      }
    },

    # Anchor date: prefer Feb 2020, else last observed pre-March 2020
    base_date_used = {
      d_feb <- date[date == base_date]
      if (length(d_feb) == 1 && !is.na(d_feb)) {
        d_feb
      } else {
        d_pre <- date[date < start_date]
        if (length(d_pre) == 0) as.Date(NA) else tail(d_pre, 1)
      }
    },

    ## --- Location-specific trend ---
    mom_chg_prepan_trnd = case_when(
      !is_post ~ mom_chg_actual,
      TRUE ~ avg_mom_growth_2017_2019_m
    ),
    growth_mult_own = case_when(
      is_post ~ 1 + avg_mom_growth_2017_2019_m,
      TRUE ~ 1
    ),
    cum_mult_own = cumprod(replace_na(growth_mult_own, 1)),
    base_cum_mult_own = {
      m0 <- cum_mult_own[date == base_date_used]
      if (length(m0) == 1 && !is.na(m0)) m0 else NA_real_
    },
    price_all_prepan_trnd = case_when(
      !is_post ~ price_all_actual,
      TRUE ~ base_price * (cum_mult_own / base_cum_mult_own)
    ),

    ## --- National-average trend (same seasonal rates for every location) ---
    mom_chg_prepan_trnd_natl = case_when(
      !is_post ~ mom_chg_actual,
      TRUE ~ avg_mom_growth_natl_m
    ),
    growth_mult_natl = case_when(
      is_post ~ 1 + avg_mom_growth_natl_m,
      TRUE ~ 1
    ),
    cum_mult_natl = cumprod(replace_na(growth_mult_natl, 1)),
    base_cum_mult_natl = {
      m0 <- cum_mult_natl[date == base_date_used]
      if (length(m0) == 1 && !is.na(m0)) m0 else NA_real_
    },
    price_all_prepan_trnd_natl = case_when(
      !is_post ~ price_all_actual,
      TRUE ~ base_price * (cum_mult_natl / base_cum_mult_natl)
    )
  ) %>%
  ungroup() %>%
  select(-population) %>%   # drop the monthly value; replaced by pop_lookup's single latest figure
  left_join(pop_lookup, by = "location_id") %>%
  rename(population = population_latest) %>%
  select(
    location_id, location_type, location_name, population, year, month,
    price_all_actual, mom_chg_actual,
    price_all_prepan_trnd, mom_chg_prepan_trnd,
    price_all_prepan_trnd_natl, mom_chg_prepan_trnd_natl
  ) %>%
  arrange(location_type, location_name, year, month)

# Quick check around the breakpoint for national
final_df %>%
  filter(location_id == 0, year %in% 2019:2021) %>%
  print(n = 50)

readr::write_csv(final_df, "final_df.csv")

readr::write_csv(final_df, "/Users/christophersalviati/Library/CloudStorage/GoogleDrive-csalviati@apartmentlist.com/Shared drives/Comms Team/Research Team/_Articles/Rent Growth vs Pre-Pandemic Trend/final_df.csv")

readr::write_csv(final_df, "/Users/christophersalviati/Library/CloudStorage/GoogleDrive-csalviati@apartmentlist.com/Shared drives/Comms Team/Research Team/_Articles/Rent Growth vs Pre-Pandemic Trend/rent_growth_v_prepandemic_trend_shiny/final_df.csv")
```
  National chart: actual vs. counterfactual trend
```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(showtext)
library(systemfonts)

library(grid)
library(gridExtra)
library(png)
library(magick)

has_ragg <- requireNamespace("ragg", quietly = TRUE)

# ---- Font setup ----
available_families <- unique(systemfonts::system_fonts()$family)
font_family <- if ("Proxima Nova" %in% available_families) "Proxima Nova" else "sans"
showtext_auto(enable = TRUE)

# ---- Build plot data ----
national_plot_df <- final_df %>%
  filter(location_id == 0) %>%
  mutate(date = ymd(sprintf("%04d-%02d-01", year, month))) %>%
  select(date, price_all_actual, price_all_prepan_trnd) %>%
  pivot_longer(
    cols = c(price_all_actual, price_all_prepan_trnd),
    names_to = "series",
    values_to = "price_all"
  ) %>%
  mutate(
    series = recode(
      series,
      price_all_actual = "Actual",
      price_all_prepan_trnd = "Pre-pandemic trend"
    ),
    series = factor(series, levels = c("Actual", "Pre-pandemic trend"))
  )

# ---- Output sizing ----
dpi <- 96

# Inner plot image size (content area) + outer padding
inner_w_px <- 754
inner_h_px <- 600
pad_px <- 5

out_w_px <- inner_w_px + 2 * pad_px
out_h_px <- inner_h_px + 2 * pad_px

# Footer + logo sizing (per your update)
footer_h_px <- 60
logo_h_px <- 40

# Title block to prevent overlap (extra space above plot area)
title_h_px <- 55  # adjust up/down if you want more/less space

# Helpers
px_to_in <- function(px, dpi) px / dpi
px_to_pt <- function(px, dpi) px * 72 / dpi

footer_h_pt <- px_to_pt(footer_h_px, dpi)
title_h_pt  <- px_to_pt(title_h_px, dpi)
pad_pt      <- px_to_pt(pad_px, dpi)
logo_h_in   <- px_to_in(logo_h_px, dpi)

# ---- Title text ----
title_line_1 <- "United States Median Rent"
title_line_2 <- "Actual vs. Pre-Pandemic Trend: 2017-Present"

# ---- Main chart (no title inside plot; we add a separate title row above) ----
p_main <- ggplot(national_plot_df, aes(x = date, y = price_all, linetype = series)) +
  geom_line(color = "#8434D6", linewidth = 1) +
  scale_linetype_manual(values = c("Actual" = "solid", "Pre-pandemic trend" = "dashed")) +
  scale_y_continuous(
    limits = c(1000, 1500),
    breaks = seq(1000, 1500, 100),
    labels = label_dollar(accuracy = 1)
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = NULL, linetype = NULL) +
  theme_minimal(base_family = font_family, base_size = 16) +
  theme(
    # Gridlines: only labeled $100 intervals; no minor gridlines
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(),
    panel.grid.major.x = element_blank(),

    # X ticks (hash marks)
    axis.ticks.x = element_line(),
    axis.ticks.length.x = unit(3, "pt"),

    # Legend in upper-left of plot area, 2 rows, clearer dash sample
    legend.position = c(0.02, 0.93),
    legend.justification = c(0, 1),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.key.height = unit(14, "pt"),
    legend.key.width  = unit(52, "pt"),  # longer sample so dash is obvious
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.margin = margin(4, 6, 4, 6),

    text = element_text(size = 16, family = font_family),
    axis.title.y = element_blank(),

    # Keep plot area tight; title is handled outside
    plot.margin = margin(t = 0, r = 6, b = 0, l = 6)
  ) +
  guides(
    linetype = guide_legend(
      nrow = 2, byrow = TRUE,
      override.aes = list(color = "#8434D6", linewidth = 1.2)
    )
  )

# ---- Title grob row (prevents overlap) ----
title_grob <- grobTree(
  textGrob(
    title_line_1,
    x = unit(0.5, "npc"), y = unit(0.72, "npc"),
    just = c("center", "center"),
    gp = gpar(fontfamily = font_family, fontsize = 20)
  ),
  textGrob(
    title_line_2,
    x = unit(0.5, "npc"), y = unit(0.28, "npc"),
    just = c("center", "center"),
    gp = gpar(fontfamily = font_family, fontsize = 16)
  )
)

# ---- Logo path (now local file in same folder as .Rmd) ----
logo_path <- "Apartment_List_Logo_RGB crop.png"

logo_path_norm <- tryCatch(
  normalizePath(logo_path, winslash = "/", mustWork = FALSE),
  error = function(e) logo_path
)

message("Working directory: ", getwd())
message("Logo exists? ", file.exists(logo_path_norm))

# ---- Pre-resize logo to EXACT 40px height (high-quality) ----
logo_small_path <- sub("\\.(png|PNG)$", "_h40.png", logo_path_norm)

if (file.exists(logo_path_norm)) {
  magick::image_read(logo_path_norm) %>%
    magick::image_resize(paste0("x", logo_h_px), filter = "Lanczos") %>%
    magick::image_write(path = logo_small_path, format = "png")
} else {
  warning("Logo file not found; skipping logo. Path tried: ", logo_path_norm)
}

# ---- Create logo grob ----
logo_grob <- NULL
if (file.exists(logo_small_path)) {
  logo_img <- png::readPNG(logo_small_path)
  logo_grob <- rasterGrob(
    logo_img,
    x = unit(1, "npc"),
    y = unit(0.5, "npc"),
    just = c("right", "center"),
    height = unit(logo_h_in, "in"),
    interpolate = TRUE
  )
}

# ---- Footer grob (source size 14) ----
footer_grob <- grobTree(
  textGrob(
    "Source: Apartment List",
    x = unit(0, "npc"),
    y = unit(0.5, "npc"),
    just = c("left", "center"),
    gp = gpar(fontfamily = font_family, fontsize = 14)
  ),
  if (!is.null(logo_grob)) logo_grob else nullGrob()
)

# ---- Assemble INNER canvas: title row + plot row + footer row ----
combined_inner <- arrangeGrob(
  title_grob,
  ggplotGrob(p_main),
  footer_grob,
  ncol = 1,
  heights = unit.c(unit(title_h_pt, "pt"), unit(1, "null"), unit(footer_h_pt, "pt"))
)

# ---- Add OUTER padding by drawing inside an "inset" viewport ----
# This guarantees visible padding in both RStudio and saved file.
padded <- grobTree(
  rectGrob(gp = gpar(col = NA, fill = "white")),  # background
  combined_inner,
  vp = viewport(
    x = unit(0.5, "npc"), y = unit(0.5, "npc"),
    width  = unit(1, "npc") - unit(2 * pad_pt, "pt"),
    height = unit(1, "npc") - unit(2 * pad_pt, "pt")
  )
)

# ---- Display in RStudio ----
grid.newpage()
grid.draw(padded)

# ---- Save PNG ----
out_file <- "national_actual_vs_prepan_trend.png"

if (has_ragg) {
  ragg::agg_png(out_file, width = out_w_px, height = out_h_px, units = "px", res = dpi)
  grid.draw(padded)
  dev.off()
} else {
  png(out_file, width = out_w_px, height = out_h_px, units = "px", res = dpi)
  grid.draw(padded)
  dev.off()
}
```
