---
title: "rent_growth_v_pre_pandemic_trend"
output: html_document
date: "2026-01-22"
---

```{r}
library(bigrquery)
library(tidyverse)
library(DBI)
```

```{r}
con <- dbConnect(
  bigrquery::bigquery(),
  project = "alist-production-scratch",
  dataset = "salviati",
  billing = "alist-production-scratch"
)
```

```{r}
index_data <- dbGetQuery(con,
  "
  select *
  from alist-production.analytics.rent_report_price_indexes;
  ") 
```
## Create alt-natl index based on weighting of metro-level
```{r}
month_ct_natl <- index_data %>% 
  filter(location_id == 0) %>%
  count() %>%
  pull()

top100_metros <- index_data %>%
  filter(location_id == cbsa_fips & !is.na(moving_avg_price_index)) %>%
  group_by(location_id, population) %>%
  summarise(month_ct = n()) %>%
  ungroup() %>%
  filter(month_ct == month_ct_natl) %>%
  top_n(100, wt = population) %>%
  select(location_id)

index_natl_alt <- index_data %>%
  filter(location_id %in% pull(top100_metros, location_id)) %>%
  group_by(year, month) %>%
  summarise(moving_avg_price_index = weighted.mean(moving_avg_price_index, wt = population),
            price_all = weighted.mean(price_all, wt = population)) %>%
  ungroup() %>%
  arrange(year, month) %>%
  mutate(yoy_chg = (moving_avg_price_index / lag(moving_avg_price_index, 12)) - 1,
         yoy_chg2 = (price_all / lag(price_all, 12)) - 1)
  

```

## Create alt-natl index based on weighting of state-level
```{r}
month_ct_natl <- index_data %>% 
  filter(location_id == 0) %>%
  count() %>%
  pull()

states <- index_data %>%
  filter(!is.na(state_fips) & is.na(cbsa_fips) & is.na(county_fips) & !is.na(moving_avg_price_index)) %>%
  group_by(location_id, population) %>%
  summarise(month_ct = n()) %>%
  ungroup() %>%
  filter(month_ct == month_ct_natl) %>%
  top_n(100, wt = population) %>%
  select(location_id)

index_natl_alt <- index_data %>%
  filter(location_id %in% pull(top100_metros, location_id)) %>%
  group_by(year, month) %>%
  summarise(moving_avg_price_index = weighted.mean(moving_avg_price_index, wt = population, na.rm = TRUE),
            price_all = weighted.mean(price_all, wt = population, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(year, month) %>%
  mutate(yoy_chg = (moving_avg_price_index / lag(moving_avg_price_index, 12)) - 1,
         yoy_chg2 = (price_all / lag(price_all, 12)) - 1)


index_natl_alt2 <- index_data %>%
  filter(location_id %in% pull(top100_metros, location_id)) %>%
  mutate(moving_avg_price_index_wt = moving_avg_price_index * population,
         price_all_wt = price_all * population) %>%
  select(year, month, moving_avg_price_index_wt, price_all_wt, population) %>%
  group_by(year, month) %>%
  summarise_all(sum) %>%
  ungroup() %>%
  arrange(year, month) %>%
  mutate(moving_avg_price_index = moving_avg_price_index_wt / population,
         price_all = price_all_wt / population,
         yoy_chg = (moving_avg_price_index / lag(moving_avg_price_index, 12)) - 1,
         yoy_chg2 = (price_all / lag(price_all, 12)) - 1,
         location_id = -1) %>%
  select(location_id, year, month, moving_avg_price_index, price_all) 
  

```


```{r}
start_date <- ymd("2020-03-01")
base_date  <- start_date %m-% months(1)  # Typically 2020-02-01

prepan_year_min <- 2017L
prepan_year_max <- 2019L

```


Identify national + top 50 metros
Metro identification logic

A row is treated as metro-level when:

cbsa_fips is not missing, and

county_fips is missing

Adjust this logic if your dataset defines metros differently.
```{r}
df_locs <- index_data %>%
  bind_rows(index_natl_alt2) %>%
  mutate(
    date = ymd(sprintf("%04d-%02d-01", year, month)),
    is_national = location_id <= 0,
    is_metro = !is.na(cbsa_fips) & is.na(county_fips)
  )

# Identify the 50 largest metros by population (using max observed population per location_id)
top_50_metros <- df_locs %>%
  filter(is_metro) %>%
  group_by(location_id) %>%
  summarise(population = max(population, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(population)) %>%
  slice_head(n = 50) %>%
  pull(location_id)

locations_keep <- c(-1, 0, top_50_metros)

df_keep <- df_locs %>%
  filter(location_id %in% locations_keep) %>%
  select(
    location_id, year, month, date,
    population, price_all, moving_avg_price_index,
    cbsa_fips, county_fips
  )

# Quick check: should be 51 locations (1 national + 50 metros)
temp <- df_keep %>%
  distinct(location_id) %>%
  summarise(n_locations = n())
```


Compute actual MoM growth rates

MoM growth is calculated using moving_avg_price_index.
```{r}
df_mom <- df_keep %>%
  arrange(location_id, date) %>%
  group_by(location_id) %>%
  mutate(
    price_all_actual = price_all,
    mom_chg_actual = (moving_avg_price_index / lag(moving_avg_price_index)) - 1
  ) %>%
  ungroup()

```

Compute average MoM growth during 2017–2019
```{r}
avg_prepan_growth_monthly <- df_mom %>%
  filter(year >= prepan_year_min, year <= prepan_year_max) %>%
  group_by(location_id, month) %>%
  summarise(
    avg_mom_growth_2017_2019_m = mean(mom_chg_actual, na.rm = TRUE),
    .groups = "drop"
  )

# Quick peek: national seasonal pattern
avg_prepan_growth_monthly %>%
  filter(location_id == 0) %>%
  arrange(month)
```

Construct the counterfactual pre-pandemic trend

Method:

Use actual prices through February 2020 (or last available pre-March observation).

From March 2020 onward, apply a constant MoM growth rate equal to the location’s 2017–2019 average.

Pre-period values match actuals exactly.

```{r}
final_df <- df_mom %>%
  left_join(avg_prepan_growth_monthly, by = c("location_id", "month")) %>%
  group_by(location_id) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    is_post = date >= start_date,

    # Anchor price: prefer Feb 2020, else last observed pre-March 2020
    base_price = {
      p_feb <- price_all_actual[date == base_date]
      if (length(p_feb) == 1 && !is.na(p_feb)) {
        p_feb
      } else {
        p_last <- price_all_actual[date < start_date]
        if (length(p_last) == 0) NA_real_ else tail(p_last, 1)
      }
    },

    # Anchor date: prefer Feb 2020, else last observed pre-March 2020
    base_date_used = {
      d_feb <- date[date == base_date]
      if (length(d_feb) == 1 && !is.na(d_feb)) {
        d_feb
      } else {
        d_pre <- date[date < start_date]
        if (length(d_pre) == 0) as.Date(NA) else tail(d_pre, 1)
      }
    },

    # For post-period months, use seasonal (month-of-year) average growth; pre-period not used for projection
    mom_chg_prepan_trnd = case_when(
      !is_post ~ mom_chg_actual,
      TRUE ~ avg_mom_growth_2017_2019_m
    ),

    # Build cumulative multiplier using seasonal growth starting March 2020 (1 elsewhere)
    growth_mult = case_when(
      is_post ~ 1 + avg_mom_growth_2017_2019_m,
      TRUE ~ 1
    ),

    cum_mult = cumprod(replace_na(growth_mult, 1)),

    # Rebase the cumulative multiplier at the anchor date (Feb 2020 if present, else last pre-March month)
    base_cum_mult = {
      m0 <- cum_mult[date == base_date_used]
      if (length(m0) == 1 && !is.na(m0)) m0 else NA_real_
    },

    # Counterfactual price: equals actual pre-period; equals base_price * (cum_mult / base_cum_mult) post-period
    price_all_prepan_trnd = case_when(
      !is_post ~ price_all_actual,
      TRUE ~ base_price * (cum_mult / base_cum_mult)
    )
  ) %>%
  ungroup() %>%
  select(
    location_id, year, month,
    price_all_actual, mom_chg_actual,
    price_all_prepan_trnd, mom_chg_prepan_trnd
  ) %>%
  arrange(location_id, year, month)

# Quick check around the breakpoint for national
final_df %>%
  filter(location_id == 0, year %in% 2019:2021) %>%
  print(n = 50)

readr::write_csv(final_df, "/Users/christophersalviati/Library/CloudStorage/GoogleDrive-csalviati@apartmentlist.com/Shared drives/Comms Team/Research Team/_Articles/Rent Growth vs Pre-Pandemic Trend/final_df.csv")

readr::write_csv(final_df, "/Users/christophersalviati/Library/CloudStorage/GoogleDrive-csalviati@apartmentlist.com/Shared drives/Comms Team/Research Team/_Articles/Rent Growth vs Pre-Pandemic Trend/rent_growth_v_prepandemic_trend_shiny/final_df.csv")
```
  National chart: actual vs. counterfactual trend
```{r}
library(tidyverse)
library(lubridate)
library(scales)
library(showtext)
library(systemfonts)

library(grid)
library(gridExtra)
library(png)
library(magick)

has_ragg <- requireNamespace("ragg", quietly = TRUE)

# ---- Font setup ----
available_families <- unique(systemfonts::system_fonts()$family)
font_family <- if ("Proxima Nova" %in% available_families) "Proxima Nova" else "sans"
showtext_auto(enable = TRUE)

# ---- Build plot data ----
national_plot_df <- final_df %>%
  #filter(location_id == -1) %>%
  filter(location_id == 0) %>%
  mutate(date = ymd(sprintf("%04d-%02d-01", year, month))) %>%
  select(date, price_all_actual, price_all_prepan_trnd) %>%
  pivot_longer(
    cols = c(price_all_actual, price_all_prepan_trnd),
    names_to = "series",
    values_to = "price_all"
  ) %>%
  mutate(
    series = recode(
      series,
      price_all_actual = "Actual",
      price_all_prepan_trnd = "Pre-pandemic trend"
    ),
    series = factor(series, levels = c("Actual", "Pre-pandemic trend"))
  )

# ---- Output sizing ----
dpi <- 96

# Inner plot image size (content area) + outer padding
inner_w_px <- 754
inner_h_px <- 600
pad_px <- 5

out_w_px <- inner_w_px + 2 * pad_px
out_h_px <- inner_h_px + 2 * pad_px

# Footer + logo sizing (per your update)
footer_h_px <- 60
logo_h_px <- 40

# Title block to prevent overlap (extra space above plot area)
title_h_px <- 55  # adjust up/down if you want more/less space

# Helpers
px_to_in <- function(px, dpi) px / dpi
px_to_pt <- function(px, dpi) px * 72 / dpi

footer_h_pt <- px_to_pt(footer_h_px, dpi)
title_h_pt  <- px_to_pt(title_h_px, dpi)
pad_pt      <- px_to_pt(pad_px, dpi)
logo_h_in   <- px_to_in(logo_h_px, dpi)

# ---- Title text ----
title_line_1 <- "United States Median Rent"
title_line_2 <- "Actual vs. Pre-Pandemic Trend: 2017-Present"

# ---- Main chart (no title inside plot; we add a separate title row above) ----
p_main <- ggplot(national_plot_df, aes(x = date, y = price_all, linetype = series)) +
  geom_line(color = "#8434D6", linewidth = 1) +
  scale_linetype_manual(values = c("Actual" = "solid", "Pre-pandemic trend" = "dashed")) +
  scale_y_continuous(
    limits = c(1000, 1500),
    breaks = seq(1000, 1500, 100),
    labels = label_dollar(accuracy = 1)
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = NULL, linetype = NULL) +
  theme_minimal(base_family = font_family, base_size = 16) +
  theme(
    # Gridlines: only labeled $100 intervals; no minor gridlines
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(),
    panel.grid.major.x = element_blank(),

    # X ticks (hash marks)
    axis.ticks.x = element_line(),
    axis.ticks.length.x = unit(3, "pt"),

    # Legend in upper-left of plot area, 2 rows, clearer dash sample
    legend.position = c(0.02, 0.93),
    legend.justification = c(0, 1),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.key.height = unit(14, "pt"),
    legend.key.width  = unit(52, "pt"),  # longer sample so dash is obvious
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.margin = margin(4, 6, 4, 6),

    text = element_text(size = 16, family = font_family),
    axis.title.y = element_blank(),

    # Keep plot area tight; title is handled outside
    plot.margin = margin(t = 0, r = 6, b = 0, l = 6)
  ) +
  guides(
    linetype = guide_legend(
      nrow = 2, byrow = TRUE,
      override.aes = list(color = "#8434D6", linewidth = 1.2)
    )
  )

# ---- Title grob row (prevents overlap) ----
title_grob <- grobTree(
  textGrob(
    title_line_1,
    x = unit(0.5, "npc"), y = unit(0.72, "npc"),
    just = c("center", "center"),
    gp = gpar(fontfamily = font_family, fontsize = 20)
  ),
  textGrob(
    title_line_2,
    x = unit(0.5, "npc"), y = unit(0.28, "npc"),
    just = c("center", "center"),
    gp = gpar(fontfamily = font_family, fontsize = 16)
  )
)

# ---- Logo path (now local file in same folder as .Rmd) ----
logo_path <- "Apartment_List_Logo_RGB crop.png"

logo_path_norm <- tryCatch(
  normalizePath(logo_path, winslash = "/", mustWork = FALSE),
  error = function(e) logo_path
)

message("Working directory: ", getwd())
message("Logo exists? ", file.exists(logo_path_norm))

# ---- Pre-resize logo to EXACT 40px height (high-quality) ----
logo_small_path <- sub("\\.(png|PNG)$", "_h40.png", logo_path_norm)

if (file.exists(logo_path_norm)) {
  magick::image_read(logo_path_norm) %>%
    magick::image_resize(paste0("x", logo_h_px), filter = "Lanczos") %>%
    magick::image_write(path = logo_small_path, format = "png")
} else {
  warning("Logo file not found; skipping logo. Path tried: ", logo_path_norm)
}

# ---- Create logo grob ----
logo_grob <- NULL
if (file.exists(logo_small_path)) {
  logo_img <- png::readPNG(logo_small_path)
  logo_grob <- rasterGrob(
    logo_img,
    x = unit(1, "npc"),
    y = unit(0.5, "npc"),
    just = c("right", "center"),
    height = unit(logo_h_in, "in"),
    interpolate = TRUE
  )
}

# ---- Footer grob (source size 14) ----
footer_grob <- grobTree(
  textGrob(
    "Source: Apartment List",
    x = unit(0, "npc"),
    y = unit(0.5, "npc"),
    just = c("left", "center"),
    gp = gpar(fontfamily = font_family, fontsize = 14)
  ),
  if (!is.null(logo_grob)) logo_grob else nullGrob()
)

# ---- Assemble INNER canvas: title row + plot row + footer row ----
combined_inner <- arrangeGrob(
  title_grob,
  ggplotGrob(p_main),
  footer_grob,
  ncol = 1,
  heights = unit.c(unit(title_h_pt, "pt"), unit(1, "null"), unit(footer_h_pt, "pt"))
)

# ---- Add OUTER padding by drawing inside an "inset" viewport ----
# This guarantees visible padding in both RStudio and saved file.
padded <- grobTree(
  rectGrob(gp = gpar(col = NA, fill = "white")),  # background
  combined_inner,
  vp = viewport(
    x = unit(0.5, "npc"), y = unit(0.5, "npc"),
    width  = unit(1, "npc") - unit(2 * pad_pt, "pt"),
    height = unit(1, "npc") - unit(2 * pad_pt, "pt")
  )
)

# ---- Display in RStudio ----
grid.newpage()
grid.draw(padded)

# ---- Save PNG ----
out_file <- "national_actual_vs_prepan_trend.png"

if (has_ragg) {
  ragg::agg_png(out_file, width = out_w_px, height = out_h_px, units = "px", res = dpi)
  grid.draw(padded)
  dev.off()
} else {
  png(out_file, width = out_w_px, height = out_h_px, units = "px", res = dpi)
  grid.draw(padded)
  dev.off()
}
```


